---
title: "ACS Script - currently untitled"
author: "STAT W4249 Cycle 1-7"
date: "January 27, 2016"
output: html_document
---

### CURRENT STATUS: Just playing around looking at characteristics of respondents who report not speaking English very well vs. not speaking English at all

Libraries needed:  
```{r message = FALSE}
library(repmis) # for reading in data from Dropbox
library(data.table)
library(dplyr)
```

repmis contains the function source_DropboxData which allows the reading of files from a Dropbox link. The
CSV files are hosted on Dropbox because Github has a file size limit of 100 MB. 
```{r}
filenames <- list(pusa_filename = "ss13pusa.csv", pusb_filename = "ss13pusb.csv",
                  husa_filename = "ss13husa.csv", husb_filename = "ss13husb.csv")
keys <- list(pusa_key = "2bag8q8jp61dsux", pusb_key = "dvvpzk7qj6r5nvb", 
             husa_key = "5uubarlztxpw947", husb_key = "xjzc0noi23zv20j")
pusa <- source_DropboxData(file = filenames$pusa_filename, key = keys$pusa_key, 
                           sep = ",", header = TRUE, cache = TRUE)
pusb <- source_DropboxData(file = filenames$pusb_filename, key = keys$pusb_key, 
                           sep = ",", header = TRUE, cache = TRUE)

# pusa and pusb have 25 states each, so we row bind them to get the complete data
pus <- rbind(pusa, pusb)
rm(pusa, pusb)

# subset the data according to the variables we're focusing on and write the subset to a new file
pus_subset <- pus %>% filter(AGEP > 5) %>% select(CIT, ENG, SCHL, ST, LANP)
rm(pus)
saveRDS(pus_subset, file = "output/pus_subset.RDS")
```

For future analyses: read the subsetted file
```{r}
above_5 <- readRDS("output/pus_subset.RDS")
```

We concentrate on people above the age of 5 (because ENG is NA for either people 5 and below OR people who speak only English, and we want to capture the latter).    

Convert CIT and ENG to factor variables and rename factor labels:
```{r}
above_5$CIT <- as.factor(above_5$CIT)
levels(above_5$CIT) <- c("Born in U.S.", "Born in U.S. territory", "Born abroad of U.S. parents", 
                     "Naturalized", "Not citizen")
plot(above_5$CIT)

above_5$ENG <- as.factor(above_5$ENG)
levels(above_5$ENG) <- c("Very well", "Well", "Not well", "Not at all", "Speaks only English")
above_5$ENG[is.na(above_5$ENG)]<- "Speaks only English"
plot(above_5$ENG)

plot(above_5$CIT[above_5$ENG == "Not well"], 
     main = "Citizenship status of those who don't speak English well")
plot(above_5$CIT[above_5$ENG == "Not at all"],
     main = "Citizenship status of those who don't speak any English")
```

