---
title: "FES map"
author: "Group 7"
date: "February 1, 2016"
output: pdf_document
---

## Load the libraries  

```{r message = FALSE}
library(mapproj)
library(dplyr)
library(gridExtra)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(choroplethr)
library(choroplethrMaps)
library(splitstackshape)
```

## Distribution by state of types of married-couple families  
Next, we look at the distribution of different types of married-couple families across the continental United States. Do certain states or groups of states have greater proportions of families with both husband and wife in the labor force, as opposed to only one of them or neither of them?  

```{r message = FALSE}
statenames <- read.csv("../data/statenames.csv", strip.white = TRUE)
statenames <- rename(statenames, ST = code, ST_name = name, ST_abbr = abbr)
hp <- readRDS("../data/hp.RDS")
hp <- tbl_df(hp)
hp <- left_join(hp, statenames)
hp_fes1to4 <- select(hp, SERIALNO, WGTP, FES, ST_name, ST_abbr) %>% 
              filter(FES >= 1 & FES <= 4) %>% 
              distinct(SERIALNO)
hp_fes1to4_wgtp <- expandRows(hp_fes1to4, "WGTP")
hp_fes1to4_state <- hp_fes1to4_wgtp %>% group_by(ST_name, FES) %>% 
                    summarise(FES_count = n()) %>% rename(region = ST_name)
hp_fes1to4_state_total <- hp_fes1to4_state %>% group_by(region) %>%
                          summarise(total_count = sum(FES_count))
total_count <- rep(hp_fes1to4_state_total$total_count, each = 4)
hp_fes1to4_state <- cbind(hp_fes1to4_state, total_count)
hp_fes1to4_state <- tbl_df(hp_fes1to4_state)
hp_fes1to4_state <- mutate(hp_fes1to4_state, value = FES_count / total_count)
hp_fes1to4_state$region <- tolower(hp_fes1to4_state$region)
hp_fes1to4_state$value <- 100 * hp_fes1to4_state$value
hp_fes1_state <- filter(hp_fes1to4_state, FES == 1)
hp_fes2_state <- filter(hp_fes1to4_state, FES == 2)
hp_fes3_state <- filter(hp_fes1to4_state, FES == 3)
hp_fes4_state <- filter(hp_fes1to4_state, FES == 4)
continental_regions <- hp_fes1_state$region[-c(2, 12)]
```

## Map  

```{r echo = FALSE, message = FALSE, fig.height = 5, fig.width = 8.5}
source("../lib/state_choropleth_wl.R")
map2 <- state_choropleth_wl(hp_fes2_state, title = "Only Husband in Labor Force",
                         num_colors = 5, zoom = continental_regions)
map2 <- map2 + scale_fill_brewer(type = "seq", palette = "Blues", name = "Percentage")
map2 <- map2 + coord_map(projection = "albers", lat0 = 29.5, lat1 = 45.5)
map3 <- state_choropleth_wl(hp_fes3_state, title = "Only Wife in Labor Force",
                         num_colors = 5, zoom = continental_regions)
map3 <- map3 + scale_fill_brewer(type = "seq", palette = "Reds", name = "Percentage")
map3 <- map3 + coord_map(projection = "albers", lat0 = 29.5, lat1 = 45.5)
map1 <- state_choropleth_wl(hp_fes1_state, title = "Both in Labor Force",
                         num_colors = 5, zoom = continental_regions)
map1 <- map1 + scale_fill_brewer(type = "seq", palette = "Purples", name = "Percentage")
map1 <- map1 + coord_map(projection = "albers", lat0 = 29.5, lat1 = 45.5)
map4 <- state_choropleth_wl(hp_fes4_state, title = "Neither in Labor Force",
                         num_colors = 5, zoom = continental_regions)
map4 <- map4 + scale_fill_grey(start = 0.8, end = 0.2, name = "Percentage")
map4 <- map4 + coord_map(projection = "albers", lat0 = 29.5, lat1 = 45.5)
map1to4 <- grid.arrange(map2, map3, map1, map4, ncol = 2, nrow = 2)
```

## What can we learn from this map?  

Across the 48 continental states:  
* The proportion of married-couple families with only the husband in the labor force varies from 14 to 30%, with evident clusters of high proportions towards the southwest and south of the country and clusters of low proportions towards the north-central part.  
* The proportion of married-couple families with only the wife in the labor force varies from 6 to 11% (in other words, it's far less common in every state for a family to have only the wife in the labor force). There appears to be clusters of high proportions around the southern states.  
* The proportion of married-couple families with both the husband and the wife in the labor force varies from 41 to 63%, with the lowest proportions clustering around the southern states and the highest proportions in the north-central part of the country.  
* The proportion of married-couple families with neither the husband or the wife in the labor force varies from 12 to 26%. We are not sure if there's any pattern here, but places like Florida with lots of retirement communities exhibit high proportions here (Florida is ranked #1 by this metric).  