---
title: "Draft Analysis for Married Family & Family Income"
author: "Yicheng Wang, Arnold Lau"
date: "January 31, 2016"
output:
  html_document:
    fig_width: 8
    number_sections: yes
    theme: united
    highlight: tango
    toc: yes
---

# Preparation
```{r, message=FALSE}
library("survey")
library("dplyr")
library("data.table")
library("ggplot2")
library("choroplethr")
library("choroplethrMaps")
```

```{r}
source("../lib/readIn.R")
hus <- readIn_house(TRUE, pathA = "../../ss13husa.csv", pathB = "../../ss13husb.csv", save = FALSE)
weights <- select(hus, WGTP)
repweights <- select(hus, starts_with("wgtp"), -WGTP)
# Specify the sampling design structure using the replicate weights
design <- svrepdesign(data = hus, repweights = repweights, weights = weights, type = "BRR", 
                      combined.weights = TRUE)
rm(weights, repweights)
```

There are a total of 1,476,313 households in the dataset. In this analysis we focus on a subset of the households via the variable __FES__ (Family Type and Employment Status), looking only at married-couple families.  

```{r}
table(hus$FES, useNA = "ifany")
sum(table(hus$FES)[1:4])
```

There are 666,947 missing values in __FES__ in the sample, and our subset will have 617,090 families. We can also get an estimate of the total number of households this represents in the entire U.S:

```{r}
hus$FES_CAT <- as.factor(hus$FES)
# Get the population estimate for the total number of households by FES
fes_popest <- as.data.frame(svytotal(~hus$FES_CAT, design = design, na.rm = TRUE))
fes_popest <- cbind(fes_popest, FES = c(1:8))
fes_popest
```

```{r}
hus <- tbl_df(hus)
hus_subset <- hus %>% select(FES, FES_CAT, FINCP, ST) %>% filter(FES %in% c(1:4)) %>% group_by(FES)
hus_subset$FES_CAT <- droplevels(hus_subset$FES_CAT)
```

# Visualization
```{r, message=F}
fescode <- "FES,MarriedFamilyEmployType
              1,Both in Labor Force
              2,Husband in Labor Force
              3,Wife in Labor Force
              4,Neither in Labor Force"
fescodes <- fread(fescode)

fesnumbers <- summarise(hus_subset,count=n())
fesnumbers <- left_join(fesnumbers,fescodes,by.x=c("FES"))
fesnumbers <- left_join(fesnumbers, fes_popest, by.x = c("FES"))
fesnumbers <- mutate(fesnumbers, samp_percent = (count / sum(fesnumbers$count)) * 100, 
                                 pop_percent = (total / sum(fesnumbers$total)) * 100)
Types <- factor(fesnumbers$MarriedFamilyEmployType,levels=unique(fesnumbers$MarriedFamilyEmployType))
```

## Descriptive Statistics

```{r}
options(scipen = 999) # display digits instead of scientific notation
ggplot(fesnumbers, aes(x = Types, y = fesnumbers$pop_percent, fill = Types)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=50, h.start = 80) +
       ylab("% of Households") + 
       xlab("Labor Force Participation") + 
       ggtitle("Labor Force Participation of Married-Couple Households in the US") +
       theme(axis.text.x = element_text(angle = 8, hjust = 0.5),
       panel.background = element_rect(fill = 'white')) +
       scale_y_discrete(breaks = seq(0, 100, 10))
```

Looking at estimated population percentages for family type and employment status among married couples, we see that just over 50% of married-couple households have both husband and wife in the labor force, compared to around 22% with only the husband in the labor force, 8% with only the wife in the labor force, and 18% with neither the husband nor the wife in the labor force, which may indicate retired couples.  

```{r}
hus_subset$FINCP_ADJ <- hus_subset$FINCP * 1.007549 # adjusts reported income to 2013 dollars
summary(hus_subset$FINCP_ADJ)
```

The median family income in the sample is $77,580. (This is different from household income, where a household might consist of multiple families or multiple unrelated people living together.) Is there a difference in labor force participation between households at or above median income vs. below? Under the unlikely theory that family income has nothing to do with married-couple labor force participation, we would expect the proportion of each category for the "at or above median" vs. "below median" subgroups to be roughly the same as the general proportion.  

```{r,message=F}
abovemedian <- hus_subset %>% filter(FINCP_ADJ >= 77580) %>% group_by(FES) %>% summarise(count = n())
abovemedian <- left_join(abovemedian, fescodes, by.x=c("FES"))
abovemedian <- mutate(abovemedian, percent = (count / sum(abovemedian$count)) * 100)

belowmedian <- hus_subset %>% filter(FINCP_ADJ < 77580) %>% group_by(FES) %>% summarise(count = n())
belowmedian <- left_join(belowmedian, fescodes, by.x=c("FES"))
belowmedian <- mutate(belowmedian, percent = (count / sum(belowmedian$count)) * 100)
```

```{r,message=TRUE}
ggplot(abovemedian, aes(x = Types, y = abovemedian$percent, fill = Types)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=50, h.start = 80) +
       geom_segment(aes(x = 0.5, y = fesnumbers$pop_percent[1], 
                        xend = 1.5, yend = fesnumbers$pop_percent[1])) +
       geom_segment(aes(x = 1.5, y = fesnumbers$pop_percent[2], 
                        xend = 2.5, yend = fesnumbers$pop_percent[2])) +
       geom_segment(aes(x = 2.5, y = fesnumbers$pop_percent[3], 
                        xend = 3.5, yend = fesnumbers$pop_percent[3])) +
       geom_segment(aes(x = 3.5, y = fesnumbers$pop_percent[4], 
                        xend = 4.5, yend = fesnumbers$pop_percent[4])) +
       ylab("Percentage") + 
       xlab("Labor Force Participation") + 
       ggtitle("Labor Force Participation of Married-Couple Families above Median Income")+
       theme(axis.text.x = element_text(angle = 8, hjust = 0.5),
       panel.background = element_rect(fill = 'white')) +
       scale_y_discrete(breaks = seq(0, 100, 10)) +
       annotate("text", x = 3.5, y = 50, label = "Black lines indicate percentages over") +
       annotate("text", x = 3.5, y = 47, label = "all married-couple families")

ggplot(belowmedian, aes(x = Types , y = belowmedian$percent, fill = Types)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=50, h.start = 80) +
       geom_segment(aes(x = 0.5, y = fesnumbers$pop_percent[1], 
                        xend = 1.5, yend = fesnumbers$pop_percent[1])) +
       geom_segment(aes(x = 1.5, y = fesnumbers$pop_percent[2], 
                        xend = 2.5, yend = fesnumbers$pop_percent[2])) +
       geom_segment(aes(x = 2.5, y = fesnumbers$pop_percent[3], 
                        xend = 3.5, yend = fesnumbers$pop_percent[3])) +
       geom_segment(aes(x = 3.5, y = fesnumbers$pop_percent[4], 
                        xend = 4.5, yend = fesnumbers$pop_percent[4])) +
       ylab("Percentage") + 
       xlab("Labor Force Participation") + 
       ggtitle("Labor Force Participation of Married-Couple Families below Median Income")+
       theme(axis.text.x = element_text(angle = 8, hjust = 0.5),
       panel.background = element_rect(fill = 'white')) +
       scale_y_discrete(breaks = seq(0, 100, 10)) +
       annotate("text", x = 3, y = 42, label = "Black lines indicate percentages over") +
       annotate("text", x = 3, y = 39, label = "all married-couple families")
```

Comparing married-couple households at or above median income and married-couple households below median income, it appears that the biggest differences are in two-income vs no-income families. The percentage of single-earner families appears to be roughly the same in both groups, while there is a greater percentage of two-income families earning above median income and a greater percentage of no-income families earning below median income.  

```{r}
statenames <- read.csv("../data/statenames.csv", strip.white = TRUE)
statenames <- rename(statenames, ST = code)
```

## Choropleth Map  
This map shows percentage of family which neither of the member is in the labor force while having more than $122100 (3rd quantile) family income. 
```{r}
stateTotalBoth <- hus_subset %>% filter(FES==4) %>% group_by(ST) %>% summarise(count=n())
medianBoth <- hus_subset %>% filter(FES==4, FINCP>= 122100)	%>% group_by(ST) %>%	summarise(count=n())
medianBoth <- left_join(medianBoth, statenames)
medianBoth[is.na(medianBoth)] <- 0
medianBoth <- mutate(medianBoth,value=medianBoth$count/stateTotalBoth$count*100)
medianBoth <- rename(medianBoth, region = name)
medianBoth$region <- tolower(medianBoth$region)
state_choropleth(medianBoth,
         title = "% Married-Couple Families with Neither in Labor Force Earning > $122,100", 
         num_colors = 3)
```




