---
title: "Draft Analysis for Married Family & Family Income"
author: "Yicheng Wang, Arnold Lau"
date: "January 31, 2016"
output:
  html_document:
    fig_width: 8
    number_sections: yes
    theme: united
    highlight: tango
    toc: yes
---

# Preparation
````{r, message=FALSE}
library("survey")
library("dplyr")
library("data.table")
library("ggplot2")
library("choroplethr")
library("choroplethrMaps")
````

```{r}
source("../lib/readIn.R")
columns <- c("ST", "FES", "FINCP", "WGTP")
hus <- readIn_house(TRUE, pathA = "../../ss13husa.csv", pathB = "../../ss13husb.csv", save = FALSE)
weights <- select(hus, WGTP)
repweights <- select(hus, starts_with("wgtp"), -WGTP)
# Specify the sampling design structure using the replicate weights
design <- svrepdesign(data = hus, repweights = repweights, weights = weights, type = "BRR", 
                      combined.weights = TRUE)
rm(weights, repweights)
```

There are a total of 1,476,313 households in the dataset. In this analysis we focus on a subset of the households via the variable __FES__ (Family Type and Employment Status), looking only at married-couple families.  

```{r}
table(hus$FES, useNA = "ifany")
sum(table(hus$FES)[1:4])
```

There are 666,947 missing values in __FES__ in the sample, and our subset will have 617,090 families. We can also get an estimate of the total number of households this represents in the entire U.S:

```{r}
hus$FES_CAT <- as.factor(hus$FES)
# Get the population estimate for the total number of households by FES
FES_CAT_total <- as.data.frame(svytotal(~hus$FES_CAT, design = design, na.rm = TRUE))
FES_CAT_total <- cbind(FES_CAT_total, FES = c(1:8))
```

```{r}
hus <- tbl_df(hus)
hus_subset <- hus %>% select(FES, FES_CAT, FINCP, ST) %>% filter(FES %in% c(1:4)) %>% group_by(FES)
hus_subset$FES_CAT <- droplevels(hus_subset$FES_CAT)
````

# Visualization
````{r, message=F}
fescode <- "FES,MarriedFamilyEmployType
              1,Both in Labor Force
              2,Husband in Labor Force
              3,Wife in Labor Force
              4,Neither in Labor Force"
fescodes <- fread(fescode)

fesnumbers <- summarise(hus_subset,count=n())
fesnumbers <- left_join(fesnumbers,fescodes,by.x=c("FES"))
fesnumbers <- left_join(fesnumbers, FES_CAT_total, by.x = c("FES"))
fesnumbers <- mutate(fesnumbers, samp_percent = (count / sum(fesnumbers$count)) * 100, 
                                 pop_percent = (total / sum(fesnumbers$total)) * 100)
Types <- factor(fesnumbers$MarriedFamilyEmployType,levels=unique(fesnumbers$MarriedFamilyEmployType))
````

## Descriptive Statistics

````{r}
options(scipen = 999) # display digits instead of scientific notation
ggplot(fesnumbers, aes(x = Types, y = fesnumbers$percent, fill = Types)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=50, h.start = 80) +
       ylab("% of Households") + 
       xlab("Labor Force Participation") + 
       ggtitle("Labor Force Participation of Married-Couple Households in the US") +
       theme(axis.text.x = element_text(angle = 8, hjust = 0.5),
       panel.background = element_rect(fill = 'white')) +
       scale_y_discrete(breaks = seq(0, 100, 10))
````

Almost 50% of married-couple households have both husband and wife in the labor force, compared to 22% with only the husband in the labor force, 8.5% with only the wife in the labor force, and 20% with neither the husband nor the wife in the labor force, which may indicate retired couples.  

````{r}
hus$FINCP_ADJ <- hus$FINCP * 1.007549 # adjusts reported income to 2013 dollars
summary(hus$FINCP_ADJ)
````

The median family income is $77,580. (This is different from household income, where a household might consist of multiple families or multiple unrelated people living together.) Is there a difference in labor force participation between households at or above median income vs. below? Under the unlikely theory that family income has nothing to do with married-couple labor force participation, we would expect the proportion of each category for the "at or above median" vs. "below median" subgroups to be roughly the same as the general proportion.  

````{r,message=F}
abovemedian <- hus %>% filter(FINCP_ADJ >= 77000) %>% group_by(FES) %>% summarise(count = n())
abovemedian <- left_join(abovemedian, fescodes, by.x=c("FES"))
abovemedian <- mutate(abovemedian, percent = (count / sum(abovemedian$count)) * 100)

belowmedian <- hus %>% filter(FINCP_ADJ < 77000) %>% group_by(FES) %>% summarise(count = n())
belowmedian <- left_join(belowmedian, fescodes, by.x=c("FES"))
belowmedian <- mutate(belowmedian, percent = (count / sum(belowmedian$count)) * 100)
````

````{r,message=TRUE}
ggplot(abovemedian, aes(x = Types , y = abovemedian$percent, fill = Types)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=50, h.start = 80) +
       geom_segment(aes(x = 0.5, y = fesnumbers$percent[1], xend = 1.5, yend = fesnumbers$percent[1])) +
       geom_segment(aes(x = 1.5, y = fesnumbers$percent[2], xend = 2.5, yend = fesnumbers$percent[2])) +
       geom_segment(aes(x = 2.5, y = fesnumbers$percent[3], xend = 3.5, yend = fesnumbers$percent[3])) +
       geom_segment(aes(x = 3.5, y = fesnumbers$percent[4], xend = 4.5, yend = fesnumbers$percent[4])) +
       ylab("Percentage") + 
       xlab("Labor Force Participation") + 
       ggtitle("Labor Force Participation of Married-Couple Families above Median Income")+
       theme(axis.text.x = element_text(angle = 8, hjust = 0.5),
       panel.background = element_rect(fill = 'white')) +
       scale_y_discrete(breaks = seq(0, 100, 10)) +
       annotate("text", x = 3.5, y = 50, label = "Black lines indicate percentages over") +
       annotate("text", x = 3.5, y = 47, label = "all married-couple families")

ggplot(belowmedian, aes(x = Types , y = belowmedian$percent, fill = Types)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=50, h.start = 80) +
       geom_segment(aes(x = 0.5, y = fesnumbers$percent[1], xend = 1.5, yend = fesnumbers$percent[1])) +
       geom_segment(aes(x = 1.5, y = fesnumbers$percent[2], xend = 2.5, yend = fesnumbers$percent[2])) +
       geom_segment(aes(x = 2.5, y = fesnumbers$percent[3], xend = 3.5, yend = fesnumbers$percent[3])) +
       geom_segment(aes(x = 3.5, y = fesnumbers$percent[4], xend = 4.5, yend = fesnumbers$percent[4])) +
       ylab("Percentage") + 
       xlab("Labor Force Participation") + 
       ggtitle("Labor Force Participation of Married-Couple Families below Median Income")+
       theme(axis.text.x = element_text(angle = 8, hjust = 0.5),
       panel.background = element_rect(fill = 'white')) +
       scale_y_discrete(breaks = seq(0, 100, 10)) +
       annotate("text", x = 3, y = 42, label = "Black lines indicate percentages over") +
       annotate("text", x = 3, y = 39, label = "all married-couple families")
````

Comparing married-couple households at or above median income and married-couple households below median income, it appears that the biggest differences are in two-income vs no-income families. The percentage of single-earner families appears to be roughly the same in both groups, while there is a greater percentage of two-income families earning above median income and a greater percentage of no-income families earning below median income.  

````{r,echo=FALSE}
stateCodeCSV = "ST,region
001,alabama
002,alaska
004,arizona
005,arkansas
006,california
008,colorado
009,connecticut
010,delaware
011,district of columbia
012,florida
013,georgia
015,hawaii
016,idaho
017,illinois
018,indiana
019,iowa
020,kansas
021,kentucky
022,louisiana
023,maine
024,maryland
025,massachusetts
026,michigan
027,minnesota
028,mississippi
029,missouri
030,montana
031,nebraska
032,nevada
033,new hampshire
034,new jersey
035,new mexico
036,new york
037,north carolina
038,north dakota
039,ohio
040,oklahoma
041,oregon
042,pennsylvania
044,rhode island
045,south carolina
046,south dakota
047,tennessee
048,texas
049,utah
050,vermont
051,virginia
053,washington
054,west virginia
055,wisconsin
056,wyoming"
stateCodes <- fread(stateCodeCSV)
````


##Mapping Tool
This map shows percentage of family which neither of the member is in the labor force while having more than 121200 dollars (3rd quantile) family income. 
````{r}

stateTotalBoth <- hus	%>%
				  filter(FES==4)		%>%
				  group_by(ST)	%>%
				  summarise(count=n())
				  

medianBoth <- hus	%>%
				filter(FES==4,FINCP>= 121200)	%>%
				group_by(ST)		%>%
				summarise(count=n())
								  
medianBoth <- right_join(medianBoth,stateCodes,by.x=c("ST"))
medianBoth[is.na(medianBoth)] <- 0
medianBoth <- mutate(medianBoth,value=medianBoth$count/stateTotalBoth$count*100)

state_choropleth(medianBoth,title="Percentage of Neither People in labor earning more than 121200",num_colors=3	)
````




